---
title: "Two-sample testing"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
---
```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

```{r}
library(tidyverse)
library(emmeans)
library(rstatix)
library(WRS2)
```
# Two-sample testing, t-test examples

## Non-parametric examples  

There are two simple tests for comparing two sample means from populations: the *Student's t-test* (parametric) and the Wilcoxon rank-sum test fro errors not normally distributed (non-parametric).  But lets start considering the assumptions for these tests, like the test for difference in the samples variances.  

## Comparing Two Variances  
Before we do a test to compare two sample means, we need to compare two variances. All we do is to divide the large variance by the small variance. If the variances are the same we will get a ratio of 1, obviously. This can be done estimating the critical value of the variance ratio, for this we use the *Fisher's F* distribution. If the test statistic is larger than the critical value (*F*), we reject the null hypothesis.  For this test the null hypothesis is assuming the the variances ration is equal to one.  In R. there is a build in function called `var.test`, and rstatix has a Levene's test. We will include some variance tests in the examples bellow.  

## Wilcoxon Rank-Sum Test  
This is a non-parametric alternative to Student's *t* test, this test can be used if the error are non-normal. R has a build-in function `wilcox.test`  (and R packages too), that make the procedure without the need to estimate tables pf critical values of the Wilcoxon rank sums.

**Example 1** The purpose of a study by Tam et al. was to investigate wheelchair maneuvering in individuals with lower-level spinal cord injury (SCI) and healthy controls (C). Subjects used a modified wheelchair to incorporate a rigid seat surface to facilitate the specified experimental measurements. Interface pressure measurement was recorded by using a high-resolution pressure-sensitive mat with a spatial resolution of four sensors per square centimeter taped on the rigid seat support. During static sitting conditions, average pressures were recorded under the ischial tuberosities (the bottom part of the pelvic bones). The data for measurements of the left ischial tuberosity (in mm Hg) for the SCI and control groups are shown in EXA_C07_S03_02.csv. We wish to know if we may conclude, on the basis of these data, that, in general, healthy subjects exhibit lower pressure than SCI subjects.

*Result* We are going to run two solutions, there is a parametric solution using R's `t.test()` in my [Statistics course page](https://fabarrios.github.io/ProbEstad/Hipotesis/Hipotesis.html), here we are going to use the rstatix for non-parametric solution and WRS2 for a robust solution.  First to use the pipeline structure in rstatix we need to wrangle the data a little.

```{r}

SCI_con <- read_csv("~/DropBox/GitHub/ProbEstad/DataSets/ch07_all/EXA_C07_S03_02.csv", show_col_types = FALSE)

SCI_con

SCI_long <- SCI_con %>% 
  pivot_longer(cols = everything(), names_to = "Group", values_to = "Pressure")

SCI_long <- SCI_long %>% 
  mutate( Group = factor(Group) )

SCI_long <- SCI_long %>% 
  mutate( Group = Group %>% fct_relevel("CONTROL","SCI") )

ggplot(SCI_long, aes(x = Group, y = Pressure, color = Group)) +
  geom_boxplot() +
  geom_jitter(height = .1, width = .1)

var.test(Pressure ~ Group, data = SCI_long)
SCI_long %>% levene_test(Pressure ~ Group, center = median)

SCL_wilcox <- SCI_long %>% wilcox_test(Pressure ~ Group, p.adjust.method = "holm", paired = FALSE, alternative = "less", conf.level = 0.95) %>%
  add_significance()

SCL_wilcox

SCI_long %>% wilcox_effsize(Pressure ~ Group)
```

this answer for the estimate falls in the non-rejection region, the two groups are **not different**.  

**Example 2** Can we conclude that patients with primary hypertension (PH), on the average, have higher total cholesterol levels than normotensive (NT) patients? This was one of the inquiries of interest for Rossi et al. In the table in EXR_C07_S03_04.csv there are total cholesterol measurements (mg/dl) for 133 PH patients and 41 NT patients. Can we conclude that PH patients have, on average, higher total cholesterol levels than NT patients? Let $\alpha = .05$.  

```{r}
Hyper <- read_csv("~/DropBox/GitHub/ProbEstad/DataSets/ch07_all/EXR_C07_S03_04.csv", show_col_types = FALSE)

Hyper
```

We have the data format in long format so we can do "data and formula", nevertheless we need to "factor" the data and the levels are fine (1, 2) making the hypertensive the 1 = HT and the normotensive 2 = NT.

```{r}
Hyper <- Hyper %>% rename(Cholesterol = Length)

Hyper_n <- Hyper %>% mutate(Group = case_when(
  Group == 1 ~ "HT",
  Group == 2 ~ "NT"
))

Hyper_n <- Hyper_n %>% 
  mutate(Group = factor(Group) )

ggplot(Hyper_n, aes(x = Group, y = Cholesterol, color = Group)) +
  geom_boxplot() +
  geom_jitter(height = .1, width = .1)
```

We start this example estimating a parametric t-test, from rstatix, to use the same data format for the non-parametric test. We initiate by estimating the `var.test()`Then we run the Wilcoxon test from rstatix. Finally, in the robust statistical estimates section we use a yuen test and other examples.

```{r}
# estimate Null hypothesis is var1/var2 = 1
var.test(Cholesterol ~ Group, data = Hyper_n)

Hyper_t_test <- Hyper_n %>% 
  t_test(Cholesterol ~ Group, p.adjust.method = "holm", paired = FALSE, alternative = "less", conf.level = 0.95) %>% 
  add_significance()

Hyper_t_test

# And the non-parametric example
Hyper_wilcox <- Hyper_n %>% 
  wilcox_test(Cholesterol ~ Group, p.adjust.method = "holm", paired = FALSE, alternative = "less", conf.level = 0.95) %>% 
  add_significance()

Hyper_wilcox

```


## Robust statisticlal Methods

Now for the **Example 1** we use a  robust statistic test for a two-sample trimmed mean test from the WRS2 package, the `yuen` test and the bootstrap version too `yuenbt`

```{r}

SCL_yuen <- yuen(Pressure ~ Group, data = SCI_long)
SCL_yuen
# for effect size
set.seed(123)

yuen.effect.ci(Pressure ~ Group, data = SCI_long)
```

For the **Example 2** we use a yuen test and other robust tests that may correct for outliers o "large" values.

```{r}
Hyper_yuen <- yuen(Cholesterol ~ Group, data = Hyper_n)
Hyper_yuen

# Effect size
# set.seed(123)
akp.effect(Cholesterol ~ Group, data = Hyper_n)
yuen.effect.ci(Cholesterol ~ Group, data = Hyper_n)

```

## Paired Comparisons  
The objective of paired comparisons tests is to eliminate a maximum number of sources of extraneous variation by making the pairs similar with respect to as many variables as possible.  Instead of performing the analysis with individual observations, paired observations run the comparison on the differences, of the two observations per subject as the variable of interest.  
**Example 3** The purpose of an investigation by Morley et al. was to evaluate the analgesic effectiveness of a daily dose of oral methadone in patients with chronic neuropathic pain syndromes. The researchers used a visual analogue scale (*0–100 mm*, higher number indicates higher pain) ratings for maximum pain intensity over the course of the day. Each subject took either 20 mg of methadone or a placebo each day for 5 days. Subjects did not know which treatment they were taking. The following table gives the mean maximum pain intensity scores for the 5 days on methadone and the 5 days on placebo. Do these data provide sufficient evidence, at the .05 level of significance, to indicate that in general the maximum pain intensity is lower on days when methadone is taken? (Data is in EXR_C07_S04_03.csv).  

```{r}
# Reading the data
MaxPain <- read_csv("~/DropBox/GitHub/ProbEstad/DataSets/ch07_all/EXR_C07_S04_03.csv", show_col_types = FALSE)

# adding a new column for the subject ID
MaxPain_new <- MaxPain %>% mutate(Subj = 1:11)

# to make the data long
MaxPain_new_long <- MaxPain_new %>%
  pivot_longer(cols = c("Plac", "Meth"), names_to = "Treatment", values_to = "MaxPain")

# Now we re-level the Treatment to Placebo and Methadone to force the order of factors
MaxPain_new_long <- MaxPain_new_long %>% mutate(Treatment %>% fct_relevel("Plac", "Meth"))

# A boxplot
MaxPain_new_long %>% ggplot(aes(x = Treatment, y = MaxPain)) +
    geom_violin() +
    geom_jitter(aes(color = Treatment), shape = 16, position = position_jitter(seed = 123)) +
    labs(title = "Responces to Methadone treatment of Pain") +
    theme_bw()
```


