---
title: "LSAdata OneWay ANOVA"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="80")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```
```{r}
# Set working directory
setwd("~/Dropbox/GitHub/ProbEstad2021")

# loading libraries
library(tidyverse)
library(car)
library(emmeans)
library(rstatix)
library(multcomp)
library(ggstatsplot)
library(scales)
```

## LDS Daniel chap 8  LSADATA
One-way ANOVA exercises with large data sets.  
Refer to the serum lipid-bound sialic acid data on 1400 subjects (LSADATA). We wish to conduct a study to determine if the measurement of serum lipid-bound sialic acid (LSA) might be of use in the detection of breast cancer. The LSA measurements (mg/dl) are for four populations of subjects: normal controls, A; patients with benign breast disease, B; patients with primary breast cancer, C; and patients with recurrent metastatic breast cancer, D. Select a simple random sample of size 10 from each population and perform an appropriate analysis to determine if we may conclude that the four population means are different.Let $ \alpha = .05$ and determine the p value.  

## data import and tibble format  

First we read the large data set (LDS) for the serum lipid-bound sialic acid data on 1400 subjects (LSADATA).  

```{r}
LSAdata <- read_csv("DataSets/ch08_all/LDS_C08_LSADATA.csv")
LSAdata
```

Now change the column names according to the LDS descriptiom: Subject_number:Col1, Controls:Col2 patients with benign breast disease (BBD):Col3, patients with primary breast cancer (PBC):Col4 patients with recurrent metastatic breast cancer (RMBC)
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
